---
title: 'Appendix: Regression Models Course Project'
author: "Tamer Koksal"
output: pdf_document
---

## 1. Variable Descriptions:


Variable Name   | Variable Description
----------------|-----------------------
mpg             | Miles per gallon (US) 
cyl             | Number of cylinders
disp            | Displacement / Engine size (in cubic inches)
hp              | Horsepower
drat            | Rear axle ratio
wt              | Weight
qsec            | Qurarter mile time
vs              | Engine shape (V-engine (0) or Straight-engine(1))
am              | Transmission type (automatic (0) or manual (1))
gear            | Number of forward gears
crab            | Number of carburetor barrels

## 2. Exploratory Analysis

```{r, message=FALSE}
str(mtcars)

library(dplyr)
mcars <- mtcars
mcars <- mutate(mcars, cyl = factor(cyl, ordered = TRUE),
                vs = factor(vs, labels = c("V", "S")),
                am = factor(am, labels = c("Auto", "Manual")))
```

## 3. Pairwise comparison (scatter plots) of variables

```{r}
pairs(mcars, panel = panel.smooth)
```

## 4. Regression of `mpg` on all of the variables

```{r}
fit <- lm(mpg ~ ., data=mcars)
summary(fit)
```

## 5. Model Selection 

```{r}
# Stepwise Backward Regression
step(fit)

fitbackward <- lm(mpg ~ am + wt + qsec, data=mcars)
summary(fitbackward)

# Check for interaction effect:
fitint <- update(fitbackward, mpg ~ am + wt + qsec + am*wt)
anova(fitbackward, fitint)
summary(fitint)
```

## 6. Regression Dignostics

```{r}
par(mfrow=c(2,2))
plot(fitint)
```

